Landau quantization of nearly degenerate bands

### 2018-09-21 paper.tex
Updates from ShareLaTeX

We begin again with the Rashba model of a rotationally-symmetric Dirac cone. To convert this Dirac point to its asymmetric variant, 
let us introduce a Zeeman coupling to an in-plane magnetic field $B_\parallel$ (parallel to the $+y$ direction):
\e{ 
H_{RZ}=\f{{\hbar^2}k^{2}}{2m}+\delta \var; \;\;\;\delta \var=\alpha(k_{x}\sigma_{y}-k_{y}\sigma_{x})+\zpar\sigma_{y}, \la{eq:RZ-Hamiltonian}}
with $\zpar{:}{=}{g_s}\mu_B B_{\parallel}/2$.
While this field breaks both time-reversal ($T$) and two-fold rotational ($C_2$) symmetries, the composition $C_2T$ is preserved as $\sigma_x H^*_{RZ}(\bk)\sigma_x{=}H_{RZ}(\bk),$ which constrains spin to lie in-plane at each $\bk$. In this symmetry class, Dirac nodes are irremovable but movable   in two-dimensional $\bk$-space. The in-plane Zeeman field has the dual effect of  moving the Dirac node  to the position:
\begin{eqnarray}
\bar{k}_x & = & -\frac{\zpar}{\alpha},\;\bar{k}_y=0,\;\epsilon_{0}=\frac{{\hbar^2}\zpar^{2}}{2m\alpha^2},\label{whereisdiracpoint}
\end{eqnarray}
and tilting the Dirac cone. To observe this tilt, consider the linearized Hamiltonian in the vicinity of this node: 
\e{
H & =\alpha\left[\left(\sigma_y-\frac{{\hbar^2}\zpar}{m\alpha^2}\right)\delta k_{x}-\delta k_{y}\sigma_{x}\right]+\epsilon_{0}.}
When $|\hbar^2\zpar|>|m\alpha^2|$, the Dirac cone tilts over in the $x$ direction, leading to a Lifshitz transition  of the constant-energy band contours -- from circular to hyperbolic (in the linearized approximation), as illustrated in Fig. \ref{fig:RZ}(c). An over-tilted Dirac cone shall be referred to as type-II. Our case study of a II-Dirac point  connects  two electron-like  pockets [cf.\ \fig{fig:RZ}(a,c)], and  differs from previous case studies [Obrien, AALG] that connect an electron- and hole-like pocket.

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{blochsphere.png}
\caption{Energy dependence of the single-band Berry phase for inner (upper panel) and outer (lower panel) orbits, denoted $\phi_-^B$ and $\phi_+^B$ respectively. $\phi_{\pm}^B$  equals half the solid angle subtended by the three-vector ${\pm} \bd(\bk)$,[Berry] where  $\bd{=}({-}\alpha k_y$,$Z_\parallel{+}\alpha k_x,{-}Z_\perp$) is obtained from expressing the effective Hamiltonian [cf.\ Eq. (\ref{calHRZ})] as $\sum_jd_j\sigma_j$.  The solid angles are enclosed by blue outlines and the energy is measured from $\epsilon_0$ [cf.\ Eq. (\ref{whereisdiracpoint})].\label{fig:blochsphere}}
\end{figure}

@@

\subsection{Codimension analysis in the absence of symmetry}\la{sec:introducecodimension}

@@

\subsection{Rotational symmetric magnetic breakdown}\la{sec:rotsymmbreakdown}

@@

@@

\begin{table}
\begin{tabular*}{\columnwidth}{l@{\extracolsep{\fill}}ccccc}
\hlineB{2.0}
             & $u$ & $s$ & quasi. & $g$ & codim.\\
\hline
$\forall \bk \in \frako_0$, $g\circ \bk=\bk$ & 0 & 0 & $\surd$ & $\breve{g}\propto I$ & 3  \\
    &  &  & & $\breve{g}\not\propto I$ & 1 \\
    & 0 & 1 & & $g^2=1$ & 1  \\
    &  &  & & $g^2=-1$ & 3  \\
$|g\circ \frako_0| = |\frako_0|$ & 0 & 0 & $\surd$ & - & 1   \\
$\exists \bk \in \frako_0, g\circ \bk\ne\bk$    & 0 & 1 & & - & 1   \\
    & 1 & 0 & & $\breve{g} \propto \sigma_z$ & 2  \\
    &  &  & & $\breve{g} \not\propto \sigma_z$ & 0  \\
    & 1 & 1 & $\surd$ & $g^2=1$ & 2  \\
    &  &  & & $g^2=-1$ & 0  \\
$|g\circ \frako_0| \ne |\frako_0|$ & 0 & 0 & $\surd$ & - & 3 \\
    & 0 & 1 & & - & 3 \\
    & 1 & 0 & & - & 3 \\
    & 1 & 1 & $\surd$ & - & 3 \\
\hlineB{2.0}
\end{tabular*}
\caption{Codimensions of degeneracy manifold of two-by-two propagator $\A$ under constraints of different type of symmetries. The first column defines how the symmetry act on the cyclotron orbit, where $g\circ \bk$ (resp. $g\circ\frako$) is the image of $\bk$ (resp. $\frako$) under the the action of $g$. $|g\circ \frako=\frako|$ indicates the image of the orbit coincides with itself, regardless of orientation of the orbit. The next two columns are intrinsic properties of $g$. $u=0$ (resp. $u=1$) indicates the symmetry is a proper (resp. improper) transformation of the plane of the cyclotron orbit ($k_z=\text{constant}$). $g$ of $u=0$ (resp. $u=1$) does not flip (resp. flips) the orientation of the cyclotron orbit. $s=0$ (resp. $s=1$) indicates the symmetry is unitary (resp. antiunitary). In magnetic groups, antiunitary symmetry operations contain time reversal symmetry. The fourth column indicates whether this symmetry class applies to quasidegenerate propagator $\A$. For the five classes not applicable here, they are useful for Landau levels of fully degenerate bands, as discussed in Sec. \ref{sec:discussion}. The fifth column specifies additional conditions (on the representation of $g$) which are necessary to obtain a unique answer for codimension. $\breve{g}$ is the diagonalized, unitary matrix representation of $g$ in the degenerate subspace. The last column is codimension of the degeneracy manifold.\label{table:codimension}}
\end{table}

>>>>

We begin again with the Rashba model of a rotationally-symmetric Dirac cone. To convert this Dirac point to its asymmetric variant, 
let us introduce a Zeeman coupling to an in-plane magnetic field $B_\parallel$ (parallel to the $+y$ direction):
\e{ 
H_{RZ}=\f{{\hbar^2}k^{2}}{2m}+\delta \var; \;\;\;\delta \var=\alpha(k_{x}\sigma_{y}-k_{y}\sigma_{x})+\zpar\sigma_{y}, \la{eq:RZ-Hamiltonian}}
with $\zpar{:}{=}{g_s}\mu_B B_{\parallel}/2$.
While this field breaks both time-reversal ($T$) and two-fold rotational ($C_2$) symmetries, the composition $C_2T$ is preserved as $\sigma_x H^*_{RZ}(\bk)\sigma_x{=}H_{RZ}(\bk),$ which constrains spin to lie in-plane at each $\bk$. In this symmetry class, Dirac nodes are irremovable but movable   in two-dimensional $\bk$-space. The in-plane Zeeman field has the dual effect of  moving the Dirac node  to the position:
\begin{eqnarray}
\bar{k}_x & = & -\frac{\zpar}{\alpha},\;\bar{k}_y=0,\;\epsilon_{0}=\frac{{\hbar^2}\zpar^{2}}{2m\alpha^2},\label{whereisdiracpoint}
\end{eqnarray}
and tilting the Dirac cone. To observe this tilt, consider the linearized Hamiltonian in the vicinity of this node: 
\e{
H & =\alpha\left[\left(\sigma_y-\frac{{\hbar^2}\zpar}{m\alpha^2}\right)\delta k_{x}-\delta k_{y}\sigma_{x}\right]+\epsilon_{0}.}
When $|\hbar^2\zpar|>|m\alpha^2|$, the Dirac cone tilts over in the $x$ direction, leading to a Lifshitz transition  of the constant-energy band contours -- from circular to hyperbolic (in the linearized approximation), as illustrated in Fig. \ref{fig:RZ}(c). An over-tilted Dirac cone shall be referred to as type-II. Our case study of a II-Dirac point  connects  two electron-like  pockets [cf.\ \fig{fig:RZ}(a,c)], and  differs from previous case studies [OBrien, AALG] that connect an electron- and hole-like pocket.

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{blochsphere.png}
\caption{Solid-angle representation of  the single-band Berry phase $\phi_{\pm}^B$ for inner (subscript $-$, upper panel) and outer ($+$, lower panel) orbit, at various energies close to the Dirac point $\epsilon_0$ [cf.\ Eq. (\ref{whereisdiracpoint})]. The decomposition of the effective Hamiltonian [cf.\ Eq. (\ref{calHRZ})] into sums of Pauli matrices ($\calh{=}\sum_jd_j\sigma_j$) defines a three-vector $\bd{=}({-}\alpha k_y$,$Z_\parallel{+}\alpha k_x,{-}Z_\perp$).
 $\phi_{\pm}^B$  equals half the solid angle subtended by   ${\pm} \bd(\bk)$ \textbf{[I wasn't  careful about minus signs here]} as $\bk$ is varied along the outer/inner orbit.[Berry]   These angles are indicated in the figure by blue outlines.\label{fig:blochsphere}}
\end{figure}

@@

\subsection{Codimension analysis without symmetry, and with band-conservation symmetry}\la{sec:introducecodimension}

@@

\subsection{Codimension reduction for 2DEG with distinct, symmetry-related II-Dirac points}\la{sec:rotsymmbreakdown}

@@

In the adiabatic regime of two quasidegenerate orbits, we have found [in \s{sec:introducecodimension}] that the codimension for $\cala{=}{\pm} I$ is unity. Here we investigate how relaxing adiabaticity -- in the presence of a discrete $N$-fold rotational symmetry $\mathfrak{c}_N$  -- changes this result. One simple way to relax adiabaticity is to incorporate Landau-Zener tunneling at II-Dirac points. Since a II-Dirac cone is tilted in a special direction, a II-Dirac point cannot be the center of a rotational symmetry; the minimal model with $\mathfrak{c}_N$ symmetry thus requires $N$ distinct but symmetry-related II-Dirac points. For $N{=}2$, we will demonstrate that the codimension for $\cala{=}I$ remains unity, in accordance with the symmetric-coordinate analysis of \q{proprashba}; the existence of Landau-level quasi-crossings (over the B axis) will be explained in terms of the destructive interference of two tunneling Feynman paths

@@

\begin{table}
\begin{tabular*}{\columnwidth}{l@{\extracolsep{\fill}}cccccc}
\hlineB{2.0}
             no. & description & $u$ & $s$ & quasi. & $g$ & codim.\\
\hline
1 &$\forall \bk \in \frako_0$, $g\circ \bk=\bk$  & 0 & 0 & $\surd$ & $\breve{g}\propto I$ & 3  \\
 &    &  &  & & $\breve{g}\not\propto I$ & 1 \\
2 &    & 0 & 1 & & $g^2=1$ & 1  \\
 &    &  &  & & $g^2=-1$ & 3  \\
3 & $|g\circ \frako_0| = |\frako_0|$ & 0 & 0 & $\surd$ & - & 1   \\
4 & $\exists \bk \in \frako_0, g\circ \bk\ne\bk$    & 0 & 1 & & - & 1   \\
5 &    & 1 & 0 & & $\breve{g} \propto \sigma_z$ & 2  \\
 &    &  &  & & $\breve{g} \not\propto \sigma_z$ & 0  \\
6 &    & 1 & 1 & $\surd$ & $g^2=1$ & 2  \\
&    &  &  & & $g^2=-1$ & 0  \\
7 & $|g\circ \frako_0| \ne |\frako_0|$ & 0 & 0 & $\surd$ & - & 3 \\
8 &    & 0 & 1 & & - & 3 \\
9 &    & 1 & 0 & & - & 3 \\
10 &    & 1 & 1 & $\surd$ & - & 3 \\
\hlineB{2.0}
\end{tabular*}
\caption{Codimensions of degeneracy manifold of two-by-two propagator $\A$ under constraints of different type of symmetries. The first column defines how the symmetry act on the cyclotron orbit, where $g\circ \bk$ (resp. $g\circ\frako$) is the image of $\bk$ (resp. $\frako$) under the the action of $g$. $|g\circ \frako=\frako|$ indicates the image of the orbit coincides with itself, regardless of orientation of the orbit. The next two columns are intrinsic properties of $g$. $u=0$ (resp. $u=1$) indicates the symmetry is a proper (resp. improper) transformation of the plane of the cyclotron orbit ($k_z=\text{constant}$). $g$ of $u=0$ (resp. $u=1$) does not flip (resp. flips) the orientation of the cyclotron orbit. $s=0$ (resp. $s=1$) indicates the symmetry is unitary (resp. antiunitary). In magnetic groups, antiunitary symmetry operations contain time reversal symmetry. The fourth column indicates whether this symmetry class applies to quasidegenerate propagator $\A$. For the five classes not applicable here, they are useful for Landau levels of fully degenerate bands, as discussed in Sec. \ref{sec:discussion}. The fifth column specifies additional conditions (on the representation of $g$) which are necessary to obtain a unique answer for codimension. $\breve{g}$ is the diagonalized, unitary matrix representation of $g$ in the degenerate subspace. The last column is codimension of the degeneracy manifold.\label{table:codimension}}
\end{table}
